Finiteness in Cubical Type Theory

### 2020-06-11 sections/finiteness-predicates.tex
linked to sections in diagram

As we make our way through each predicate, we will be interested in two aspects:
how can we build proofs of this predicate (i.e. is the product of two finite
types finite?) and what do we \emph{get} once we do (i.e. does this predicate
tell us the number of elements in the finite set?).
\section{Split Enumerability}
We will start with a simple notion of finiteness, called \emph{split}
enumerability.

@@

\section{Manifest Bishop Finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst} \smallskip
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

This gets us all of the necessary closure proofs on \(\mathcal{C}\).
\section{Manifest Enumerability}
We have now explored quite far in the ``less informative'' direction.
However, all three predicates we have examined are equally \emph{restrictive}:
in this section we will see a predicate which is much less restrictive.
In particular, this predicate ranges over non-set types.

@@

  We do this using the previous recompute lemma: that tells us that all we need
  to construct is a decision for \(x \in \mathit{xs}\), and it will be able to
  derive the proof itself.
  Such a decision procedure is not difficult to construct: for any value \(x\)
  and list \(\mathit{xs}\), we proceed through the list \(\mathit{xs}\), testing
  if \(x\) is equal to any of its contents.
  If it is, we return that we have proven the goal, and that \(x\) is indeed
  present in \(\mathit{xs}\).
  Otherwise, we know that \(x\) cannot be in \(\mathit{xs}\) (since we've tested
  every value), so we return that the goal has been disproven.
\end{proof}
\section{Kuratowski Finiteness}
The one big missing definition of finiteness to cover is \emph{Kuratowski}
finiteness.
While it's quite important, it's also quite different from the definitions we've
seen so far.
It starts with an encoding of the free join semilattice.
\begin{definition}[Free Join Semilattice]
  \(\mathcal{K}(A)\) is the free join semilattice, or, alternatively, the type
  of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

>>>>

As we make our way through each predicate, we will be interested in two aspects:
how can we build proofs of this predicate (i.e. is the product of two finite
types finite?) and what do we \emph{get} once we do (i.e. does this predicate
tell us the number of elements in the finite set?).
\section{Split Enumerability} \label{split-enumerability}
We will start with a simple notion of finiteness, called \emph{split}
enumerability.

@@

\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness} \label{cardinal-finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst} \smallskip
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

This gets us all of the necessary closure proofs on \(\mathcal{C}\).
\section{Manifest Enumerability} \label{manifest-enumerability}
We have now explored quite far in the ``less informative'' direction.
However, all three predicates we have examined are equally \emph{restrictive}:
in this section we will see a predicate which is much less restrictive.
In particular, this predicate ranges over non-set types.

@@

  We do this using the previous recompute lemma: that tells us that all we need
  to construct is a decision for \(x \in \mathit{xs}\), and it will be able to
  derive the proof itself.
  Such a decision procedure is not difficult to construct: for any value \(x\)
  and list \(\mathit{xs}\), we proceed through the list \(\mathit{xs}\), testing
  if \(x\) is equal to any of its contents.
  If it is, we return that we have proven the goal, and that \(x\) is indeed
  present in \(\mathit{xs}\).
  Otherwise, we know that \(x\) cannot be in \(\mathit{xs}\) (since we've tested
  every value), so we return that the goal has been disproven.
\end{proof}
\section{Kuratowski Finiteness} \label{kuratowski}
The one big missing definition of finiteness to cover is \emph{Kuratowski}
finiteness.
While it's quite important, it's also quite different from the definitions we've
seen so far.
It starts with an encoding of the free join semilattice.
\begin{definition}[Free Join Semilattice]
  \(\mathcal{K}(A)\) is the free join semilattice, or, alternatively, the type
  of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

### 2020-06-11 sections/introduction.tex
now compiles without errors
cleaner format
linked to sections in diagram

\chapter{Introduction}
\section{Foreword}
Proofs in constructive mathematics are often more substantial than their
classical counterparts.
``Substantial'' here doesn't mean the literal length of the proofs (although the
double meaning is not lost on anyone who has had to struggle with a modern
constructive proof assistant), but rather refers to their \emph{content}: a
constructive proof contains more \emph{stuff} than a classical one.

If we were to prove that a set was finite in classical mathematics, we might
show that it is in bijection with another finite set, say a finite prefix of the
natural numbers.
In semi-formal notation, to say that a set \(A\) is finite we would show the
following:
\begin{equation} \label{example-finite}
  \mathbf{Finite}(A) \coloneqq \exists (n : \mathbb{N}) . \; A \Leftrightarrow \left\{ m \mid m < n\right\}
\end{equation}
Where \(A~\Leftrightarrow~B\) means ``a bijection between \(A\) and \(B\)''.

Classically, a proof of equation~\ref{example-finite} is nothing more than
\emph{evidence} for the proposition of equation~\ref{example-finite}.
A constructive proof of \ref{example-finite}, however, is a mathematical object
in its own right, containing:
\begin{enumerate}
  \item A number \(n\) representing the cardinality of \(A\).
  \item A function from \(A\) to \(\left\{ m \mid m < n \right\}\).
  \item A function from \(\left\{ m \mid m < n \right\}\) to \(A\).
\end{enumerate}
In other words, it makes sense (in constructive mathematics) to talk about a
function of type \(\mathbf{Finite}(A)~\rightarrow~\mathbb{N}\), which takes a
\emph{proof} of the finiteness of \(A\) and returns \(A\)'s cardinality.

Such functions are perhaps even more concrete than this description suggests:
many constructive proofs are in fact computer programs in disguise (and
vice-versa \cite{wadlerPropositionsTypes2015}).
All ``proofs'' in this paper correspond to honest-to-goodness proofs in a
programming language (Cubical Agda \cite{vezzosiCubicalAgdaDependently2019} in
our case) which can be run on a computer to produce real output.

We have now hopefully established, albeit informally, what it means for
constructive proofs to be ``substantial''.
Clearly this extra information carried by proofs can add richness to their
study: we will see, for instance, no fewer than five finiteness predicates in
this paper which would collapse to one in a classical setting.
The extra information also presents a challenge, however: occasionally we will
see that we have revealed \emph{too much} information while proving some
property.

Consider the category of finite sets.
We will explore this category in much more detail later, but for now we just
need to know that the objects of this category are---shockingly---finite sets.
Constructively, the objects of this category can be thought of as \emph{pairs},
where the first element is a set (like \(\left\{ \text{True}, \text{False}
\right\}\), the two-element set of the booleans) and the second is a proof that
that set is finite.
If we take those proofs to be the kind given in equation~\ref{example-finite},
we should see that they contain a bijection from the booleans to the set
\(\left\{ 0, 1 \right\}\): however \emph{two} such bijections exist!
\begin{figure}[h]
  \centering
  \begin{subfigure}[b]{.45\textwidth}
    \centering
    \begin{tikzcd}
      \text{True}  \ar[rr, leftrightarrow] & & 0 \\
      \text{False} \ar[rr, leftrightarrow] & & 1
    \end{tikzcd}
  \end{subfigure}
  \begin{subfigure}[b]{.45\textwidth}
    \centering
    \begin{tikzcd}
      \text{True}  \ar[rrd, leftrightarrow, start anchor=south east, end anchor=north west] & & 0 \\
      \text{False} \ar[rru, leftrightarrow, start anchor=north east, end anchor=south west, crossing over] & & 1
    \end{tikzcd}
  \end{subfigure}
  \caption{The Two Bijections Between the Booleans and \(\left\{ 0, 1 \right\}\)}
\end{figure}
This means that where we wanted one object to correspond to the finite set
\(\left\{ \text{True}, \text{False} \right\}\), we instead have two.
In fact, for any finite set of cardinality \(n\) we will have \(n!\) finiteness
proofs, and therefore \(n!\) objects.
We have constructed, in other words, entirely the wrong category.

Being able to hide the extra information in a constructive proof when we need to
is clearly important.
HoTT (and CuTT by extension) facilitate this hiding: how exactly it is
accomplished will be explored in the subsequent sections.

Finally, for all of its benefits, constructive mathematics (especially
dependently-typed constructive mathematics on a computer) still suffers from
many painful drawbacks.
As already hinted at, proofs in a constructive setting can tend to be verbose,
especially when the arbiter of correctness is as strict as a computer, which
neither understands nor trusts the time-honoured technique of ``the other cases
are obviously true''.
While the goal of using a computer to automate the boring parts of a proof still
seems attainable, too often getting a computer involved in a proof means adding
thousands of lines of tedium to satisfy its lack of intuition.

Simple proof search over a finite domain is a simple, if effective, way to
relieve some of the tedium.
Many libraries in Agda, Coq, and other languages make use of this fact.
While the idea for finite proof search is simple, we found that implementing one
revealed a number of interesting theoretical considerations:
\begin{description}
  \item[How should we construct search spaces?]
    We found that a \(Pi\)-pretopos provided the appropriate ``generator
    language'' for constructing search spaces for finite types.
  \item[How do we deal with functions?]
    We of course would like to be able to search over functions: \todo{?}
  \item[How do we talk about ``searchability''?]
    \todo{Omniscience}
\end{description}
Finiteness is a prime example of a phenomenon which happens quite frequently in
dependently-typed constructive mathematics: through building a practical program
we unearth theoretically interesting results.
\todo{rephrase}
\section{Contributions}
We are interested in constructive notions of finiteness.
In contrast to classical finiteness, in a constructive setting there is a wide
variety of predicates which all have some claim to being the formal
interpretation of ``finiteness''.
Broadly, they vary along two axes: informativeness (i.e. how many properties
of the underlying type we can derive from a proof that it is finite), and
restrictiveness (i.e. how many types conform to its particular variety of
finiteness).
We will explore five of these predicates, proving formally their relations and
implications.

@@

The arrows in figure~\ref{finite-classification} are implication functions
between each predicate.
In section~\ref{relations} we will provide a function which inhabits each arrow.
Each unlabelled arrow is an unconditional implication: every manifest Bishop
finite set is cardinal finite (lemma~\ref{manifest-bishop-to-cardinal}), for
instance.
The labelled arrows are strengthening proofs: every manifest enumerable set
\emph{with decidable equality} is split enumerable
(lemma~\ref{manifest-enum-to-split-enum}).
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.

>>>>

\chapter{Introduction}
\section{Foreword}
\todo[inline]{Foreword}
\section{Contributions}
We are interested in constructive notions of finiteness, formalised in Cubical
Type Theory.
In contrast to classical finiteness, in a constructive setting there is a wide
variety of predicates which all have some claim to being the formal
interpretation of ``finiteness''.
Broadly, they vary along two axes: informativeness (i.e. how many properties
of the underlying type we can derive from a proof that it is finite), and
restrictiveness (i.e. how many types conform to its particular variety of
finiteness).
We will explore five of these predicates, proving formally their relations and
implications.

@@

The arrows in figure~\ref{finite-classification} are implication functions
between each predicate.
We will provide a function which inhabits each arrow.
Each unlabelled arrow is an unconditional implication: every manifest Bishop
finite set is cardinal finite (lemma~\ref{manifest-bishop-to-cardinal}), for
instance.
The labelled arrows are strengthening proofs: every manifest enumerable set
\emph{with decidable equality} is split enumerable
(lemma~\ref{manifest-enum-to-split-enum}).
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.
